Study guide: Analysis of exponential decay models

### 2015-10-10 doc/src/chapters/analysis/slides_analysis.tex

% LaTeX Beamer file automatically generated from DocOnce
% https://github.com/hplgit/doconce

%-------------------- begin beamer-specific preamble ----------------------

\documentclass{beamer}

\usetheme{red_shadow}
\usecolortheme{default}

% turn off the almost invisible, yet disturbing, navigation symbols:
\setbeamertemplate{navigation symbols}{}

% Examples on customization:
%\usecolortheme[named=RawSienna]{structure}
%\usetheme[height=7mm]{Rochester}
%\setbeamerfont{frametitle}{family=\rmfamily,shape=\itshape}
%\setbeamertemplate{items}[ball]
%\setbeamertemplate{blocks}[rounded][shadow=true]
%\useoutertheme{infolines}
%
%\usefonttheme{}
%\useinntertheme{}
%
%\setbeameroption{show notes}
%\setbeameroption{show notes on second screen=right}

% fine for B/W printing:
%\usecolortheme{seahorse}

\usepackage{pgf}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{relsize}

\usepackage{fancybox}  % make sure fancybox is loaded before fancyvrb

\usepackage{fancyvrb}
\usepackage{minted} % requires pygments and latex -shell-escape filename
%\usepackage{anslistings}
%\usepackage{listingsutf8}

\usepackage{amsmath,amssymb,bm}
%\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{colortbl}
\usepackage[english]{babel}
\usepackage{tikz}
\usepackage{framed}
% Use some nice templates
\beamertemplatetransparentcovereddynamic

% --- begin table of contents based on sections ---
% Delete this, if you do not want the table of contents to pop up at
% the beginning of each section:
% (Only section headings can enter the table of contents in Beamer
% slides generated from DocOnce source, while subsections are used
% for the title in ordinary slides.)
\AtBeginSection[]
{
  \begin{frame}<beamer>[plain]
  \frametitle{}
  %\frametitle{Outline}
  \tableofcontents[currentsection]
  \end{frame}
}
% --- end table of contents based on sections ---

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\newcommand{\shortinlinecomment}[3]{\note{\textbf{#1}: #2}}
\newcommand{\longinlinecomment}[3]{\shortinlinecomment{#1}{#2}{#3}}

\definecolor{linkcolor}{rgb}{0,0,0.4}
\hypersetup{
    colorlinks=true,
    linkcolor=linkcolor,
    urlcolor=linkcolor,
    pdfmenubar=true,
    pdftoolbar=true,
    bookmarksdepth=3
    }
\setlength{\parskip}{7pt}  % {1em}

\newenvironment{doconceexercise}{}{}
\newcounter{doconceexercisecounter}
\newenvironment{doconce:movie}{}{}
\newcounter{doconce:movie:counter}

\newcommand{\subex}[1]{\noindent\textbf{#1}}  % for subexercises: a), b), etc

%-------------------- end beamer-specific preamble ----------------------

% Add user's preamble

@@

\title{Study guide: Analysis of exponential decay models}

@@

\author{Hans Petter Langtangen\inst{1,2}}
\institute{Center for Biomedical Computing, Simula Research Laboratory\inst{1}
\and
Department of Informatics, University of Oslo\inst{2}}
% ----------------- end author(s) -------------------------

\date{Oct 10, 2015
% <optional titlepage figure>
% <optional copyright>
}

\begin{frame}[plain,fragile]
\titlepage
\end{frame}

\section{Analysis of finite difference equations}

\begin{frame}[plain,fragile]
\frametitle{Analysis of finite difference equations}

\label{decay:analysis}

@@

\begin{block}{Problem setting }
How good is this method? Is it safe to use it?
\end{block}

\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Encouraging numerical solutions}

@@

% inline figure
\centerline{\includegraphics[width=1.1\linewidth]{fig-analysis/BE4c.pdf}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Discouraging numerical solutions; Crank-Nicolson}

% inline figure
\centerline{\includegraphics[width=1.1\linewidth]{fig-analysis/CN4c.pdf}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Discouraging numerical solutions; Forward Euler}

% inline figure
\centerline{\includegraphics[width=1.1\linewidth]{fig-analysis/FE4c.pdf}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Summary of observations}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Problem setting}

\begin{block}{Goal }
We ask the question

@@

\noindent
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Experimental investigation of oscillatory solutions}

@@

Seems that $a\Delta t < 1$ for FE and 2 for CN.
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Exact numerical solution}

@@

Such an exact discrete solution is unusual, but very handy for analysis.
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Stability}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Computation of stability in this problem}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Explanation of problems with Forward Euler}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Explanation of problems with Crank-Nicolson}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Summary of stability}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Comparing amplification factors}

@@

A possible measure of accuracy: $\Aex - A$
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Plot of amplification factors}

% inline figure
\centerline{\includegraphics[width=0.9\linewidth]{fig-analysis/A_factors.pdf}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{$p=a\Delta t$ is the important parameter for numerical performance}

@@

\noindent
\begin{block}{Another ``proof'' why $p=a\Delta t$ is key }
If we scale the model
by $\bar t=at$, $\bar u=u/I$, we get
$d\bar u/d\bar t = -\bar u$, $\bar u(0)=1$ (no physical parameters!).
The analysis show that $\Delta \bar t$ is key, corresponding to
$a\Delta t$ in the unscaled model.
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Series expansion of amplification factors}

@@

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{ipy}
>>> from sympy import *
>>> # Create p as a mathematical symbol with name 'p'
>>> p = Symbol('p')
>>> # Create a mathematical expression with p
>>> A_e = exp(-p)
>>>
>>> # Find the first 6 terms of the Taylor series of A_e
>>> A_e.series(p, 0, 6)
1 + (1/2)*p**2 - p - 1/6*p**3 - 1/120*p**5 + (1/24)*p**4 + O(p**6)

>>> theta = Symbol('theta')
>>> A = (1-(1-theta)*p)/(1+theta*p)
>>> FE = A_e.series(p, 0, 4) - A.subs(theta, 0).series(p, 0, 4)
>>> BE = A_e.series(p, 0, 4) - A.subs(theta, 1).series(p, 0, 4)
>>> half = Rational(1,2)  # exact fraction 1/2
>>> CN = A_e.series(p, 0, 4) - A.subs(theta, half).series(p, 0, 4)
>>> FE
(1/2)*p**2 - 1/6*p**3 + O(p**4)
>>> BE
-1/2*p**2 + (5/6)*p**3 + O(p**4)
>>> CN
(1/12)*p**3 + O(p**4)
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Error in amplification factors}

@@

\begin{equation}
A-\Aex = \left\lbrace\begin{array}{ll}
\Oof{\Delta t^2}, & \hbox{Forward and Backward Euler},\\ 
\Oof{\Delta t^3}, & \hbox{Crank-Nicolson}
\end{array}\right.
\end{equation}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The fraction of numerical and exact amplification factors}

@@

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{ipy}
>>> FE = 1 - (A.subs(theta, 0)/A_e).series(p, 0, 4)
>>> BE = 1 - (A.subs(theta, 1)/A_e).series(p, 0, 4)
>>> CN = 1 - (A.subs(theta, half)/A_e).series(p, 0, 4)
>>> FE
(1/2)*p**2 + (1/3)*p**3 + O(p**4)
>>> BE
-1/2*p**2 + (1/3)*p**3 + O(p**4)
>>> CN
(1/12)*p**3 + O(p**4)
\end{minted}
Same leading-order terms as for the error measure $A-\Aex$.
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The true/global error at a point}

\label{decay:analysis:gobal:error}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Computing the global error at a point}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{ipy}
>>> n = Symbol('n')
>>> u_e = exp(-p*n)   # I=1
>>> u_n = A**n        # I=1
>>> FE = u_e.series(p, 0, 4) - u_n.subs(theta, 0).series(p, 0, 4)
>>> BE = u_e.series(p, 0, 4) - u_n.subs(theta, 1).series(p, 0, 4)
>>> CN = u_e.series(p, 0, 4) - u_n.subs(theta, half).series(p, 0, 4)
>>> FE
(1/2)*n*p**2 - 1/2*n**2*p**3 + (1/3)*n*p**3 + O(p**4)
>>> BE
(1/2)*n**2*p**3 - 1/2*n*p**2 + (1/3)*n*p**3 + O(p**4)
>>> CN
(1/12)*n*p**3 + O(p**4)
\end{minted}
Substitute $n$ by $t/\Delta t$:

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Convergence}

The numerical scheme is convergent if the global error
$e^n\rightarrow 0$ as $\Delta t\rightarrow 0$.
If the error has a leading order term $\Delta t^r$, the
convergence rate is of order $r$.
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Integrated errors}

@@

\begin{block}{Summary of errors }
Analysis of both the pointwise and the time-integrated true errors:

@@

\noindent
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Truncation error}

@@

\begin{equation}
\frac{\uex(t_{n+1})-\uex(t_n)}{\Delta t} + a\uex(t_n) = R^n \neq 0
\label{decay:analysis:trunc:Req}
\end{equation}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Computation of the truncation error}

@@

\[ R^n \approx \half\uex''(t_n)\Delta t\]
This is a mathematical expression for the truncation error.
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The truncation error for other schemes}

@@

\[ R^{n+\half} \approx \frac{1}{24}\uex'''(t_{n+\half})\Delta t^2\]
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Consistency, stability, and convergence}

@@

(Consistency and stability is in most problems
much easier to establish than
convergence.)
\end{frame}

\end{document}

>>>>

%%
%% Automatically generated file from DocOnce source
%% (https://github.com/hplgit/doconce/)
%%

% #define PREAMBLE

% #ifdef PREAMBLE
%-------------------- begin preamble ----------------------

\documentclass[%
oneside,                 % oneside: electronic viewing, twoside: printing
final,                   % draft: marks overfull hboxes, figures with paths
10pt]{article}

\listfiles               % print all files needed to compile this document

\usepackage{relsize,makeidx,color,setspace,amsmath,amsfonts,amssymb}
\usepackage[table]{xcolor}
\usepackage{bm,microtype}

\usepackage[pdftex]{graphicx}

% Packages for typesetting blocks of computer code
\usepackage{fancyvrb,framed,moreverb}

% Define colors
\definecolor{orange}{cmyk}{0,0.4,0.8,0.2}
\definecolor{tucorange}{rgb}{1.0,0.64,0}
\definecolor{darkorange}{rgb}{.71,0.21,0.01}
\definecolor{darkgreen}{rgb}{.12,.54,.11}
\definecolor{myteal}{rgb}{.26, .44, .56}
\definecolor{gray}{gray}{0.45}
\definecolor{mediumgray}{gray}{.8}
\definecolor{lightgray}{gray}{.95}
\definecolor{brown}{rgb}{0.54,0.27,0.07}
\definecolor{purple}{rgb}{0.5,0.0,0.5}
\definecolor{darkgray}{gray}{0.25}
\definecolor{darkblue}{rgb}{0,0.08,0.45}
\definecolor{darkblue2}{rgb}{0,0,0.8}
\definecolor{lightred}{rgb}{1.0,0.39,0.28}
\definecolor{lightgreen}{rgb}{0.48,0.99,0.0}
\definecolor{lightblue}{rgb}{0.53,0.81,0.92}
\definecolor{lightblue2}{rgb}{0.3,0.3,1.0}
\definecolor{lightpurple}{rgb}{0.87,0.63,0.87}
\definecolor{lightcyan}{rgb}{0.5,1.0,0.83}

\colorlet{comment_green}{green!50!black}
\colorlet{string_red}{red!60!black}
\colorlet{keyword_pink}{magenta!70!black}
\colorlet{indendifier_green}{green!70!white}

% Backgrounds for code
\definecolor{cbg_gray}{rgb}{.95, .95, .95}
\definecolor{bar_gray}{rgb}{.92, .92, .92}

\definecolor{cbg_yellowgray}{rgb}{.95, .95, .85}
\definecolor{bar_yellowgray}{rgb}{.95, .95, .65}

\colorlet{cbg_yellow2}{yellow!10}
\colorlet{bar_yellow2}{yellow!20}

\definecolor{cbg_yellow1}{rgb}{.98, .98, 0.8}
\definecolor{bar_yellow1}{rgb}{.98, .98, 0.4}

\definecolor{cbg_red1}{rgb}{1, 0.85, 0.85}
\definecolor{bar_red1}{rgb}{1, 0.75, 0.85}

\definecolor{cbg_blue1}{rgb}{0.87843, 0.95686, 1.0}
\definecolor{bar_blue1}{rgb}{0.7,     0.95686, 1}

%\setlength{\fboxsep}{-1.5mm}  % adjust cod_vpad/pro_vpad background box

%% Background for code blocks (parameter is color name)

%% pro/cod_vpad: gives some vertical padding before and after the text
%% (but has more simplistic code than _cod/pro_tight+cod/pro).
%% pro/cod_vpad can be used to enclose Verbatim or lst begin/end for code.
%% pro/cod calls _pro/cod_tight and has very little vertical padding,
%% used to enclose Verbatim and other begin/end for code.
%% (pro/cod is what the ptex2tex program could produce with the
%% Blue/BlueBar definitions in .ptex2tex.cfg.)

\newenvironment{cod_vpad}[1]{
   \def\FrameCommand{\colorbox{#1}}
   \MakeFramed{\FrameRestore}}
   {\endMakeFramed}

\newenvironment{_cod_tight}[1]{
   \def\FrameCommand{\colorbox{#1}}
   \FrameRule0.6pt\MakeFramed {\FrameRestore}\vskip3mm}
   {\vskip0mm\endMakeFramed}

\newenvironment{cod}[1]{
\bgroup\rmfamily
\fboxsep=0mm\relax
\begin{_cod_tight}{#1}
\list{}{\parsep=-2mm\parskip=0mm\topsep=0pt\leftmargin=2mm
\rightmargin=2\leftmargin\leftmargin=4pt\relax}
\item\relax}
{\endlist\end{_cod_tight}\egroup}

%% Background for complete program blocks (parameter 1 is color name
%% for background, parameter 2 is color for left bar)
\newenvironment{pro_vpad}[2]{
   \def\FrameCommand{\color{#2}\vrule width 1mm\normalcolor\colorbox{#1}}
   \MakeFramed{\FrameRestore}}
   {\endMakeFramed}

\newenvironment{_pro_tight}[2]{
   \def\FrameCommand{\color{#2}\vrule width 1mm\normalcolor\colorbox{#1}}
   \FrameRule0.6pt\MakeFramed {\advance\hsize-2mm\FrameRestore}\vskip3mm}
   {\vskip0mm\endMakeFramed}

\newenvironment{pro}[2]{
\bgroup\rmfamily
\fboxsep=0mm\relax
\begin{_pro_tight}{#1}{#2}
\list{}{\parsep=-2mm\parskip=0mm\topsep=0pt\leftmargin=2mm
\rightmargin=2\leftmargin\leftmargin=4pt\relax}
\item\relax}
{\endlist\end{_pro_tight}\egroup}

\usepackage{listingsutf8}

% Common lstlisting parameters
\lstset{
  basicstyle=\small \ttfamily,
  breaklines=false,          % break/wrap lines
  breakatwhitespace=true,    % let linebreaks happen at whitespace
  breakindent=40pt,
  tab=,
  tabsize=4,                 % tab means 4 spaces
  %belowskip=\smallskipamount,  % space between code and text below
  xleftmargin=5pt,           % indentation of code frame
  xrightmargin=5pt,
  framexleftmargin=5pt,      % add frame space to the left of code
  %numbers=left,             % put line numbers on the left
  %stepnumber=2,             % stepnumber=1 numbers each line, =n every n lines
  %framerule=0.4pt           % thickness of frame
  aboveskip=2ex,             % vertical space above code frame
  showstringspaces=false,    % show spaces in strings with an underscore
  showspaces=false,          % show spaces with an underscore
  showtabs=false,
  keepspaces=true,
  columns=fullflexible,      % tighter character kerning, like verb
  escapeinside={||},         % for |\pause| in slides and math in code blocks
  extendedchars=\true,       % allows non-ascii chars, does not work with utf-8
}

% Internally defined styles for lstlisting

\lstdefinestyle{simple}{
commentstyle={},
}

% end of custom lstdefinestyles

\usepackage[T1]{fontenc}
%\usepackage[latin1]{inputenc}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}

\usepackage{lmodern}         % Latin Modern fonts derived from Computer Modern

% Hyperlinks in PDF:
\definecolor{linkcolor}{rgb}{0,0,0.4}
\usepackage{hyperref}
\hypersetup{
    breaklinks=true,
    colorlinks=true,
    linkcolor=linkcolor,
    urlcolor=linkcolor,
    citecolor=black,
    filecolor=black,
    %filecolor=blue,
    pdfmenubar=true,
    pdftoolbar=true,
    bookmarksdepth=3   % Uncomment (and tweak) for PDF bookmarks with more levels than the TOC
    }
%\hyperbaseurl{}   % hyperlinks are relative to this root

\setcounter{tocdepth}{2}  % number chapter, section, subsection

% Tricks for having figures close to where they are defined:
% 1. define less restrictive rules for where to put figures
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}
\renewcommand{\textfraction}{0}
\renewcommand{\floatpagefraction}{0.75}
% floatpagefraction must always be less than topfraction!
% 2. ensure all figures are flushed before next section
\usepackage[section]{placeins}
% 3. enable begin{figure}[H] (often leads to ugly pagebreaks)
%\usepackage{float}\restylefloat{figure}

\usepackage[framemethod=TikZ]{mdframed}

% --- begin definitions of admonition environments ---

% --- end of definitions of admonition environments ---

% prevent orhpans and widows
\clubpenalty = 10000
\widowpenalty = 10000

% --- end of standard preamble for documents ---

@@

\thispagestyle{empty}

\begin{center}
{\LARGE\bf
\begin{spacing}{1.25}
Study guide: Analysis of exponential decay models
\end{spacing}
}
\end{center}

@@

\begin{center}
{\bf Hans Petter Langtangen${}^{1, 2}$} \\ [0mm]
\end{center}

\begin{center}
% List of all institutions:
\centerline{{\small ${}^1$Center for Biomedical Computing, Simula Research Laboratory}}
\centerline{{\small ${}^2$Department of Informatics, University of Oslo}}
\end{center}

% ----------------- end author(s) -------------------------

% --- begin date ---
\begin{center}
Oct 10, 2015
\end{center}
% --- end date ---

\vspace{1cm}

% !split
\section*{Analysis of finite difference equations}
\label{decay:analysis}

@@

% --- begin paragraph admon ---
\paragraph{Problem setting.}
How good is this method? Is it safe to use it?
% --- end paragraph admon ---

% !split
\subsection*{Encouraging numerical solutions}

@@

% inline figure
\centerline{\includegraphics[width=1.1\linewidth]{fig-analysis/BE4c.pdf}}

% !split
\subsection*{Discouraging numerical solutions; Crank-Nicolson}

% !split
\subsection*{Discouraging numerical solutions; Forward Euler}

% !split
\subsection*{Summary of observations}

@@

\noindent
% !split
\subsection*{Problem setting}

% --- begin paragraph admon ---
\paragraph{Goal.}
We ask the question

@@

\noindent
% --- end paragraph admon ---

% !split
\subsection*{Experimental investigation of oscillatory solutions}

@@

Seems that $a\Delta t < 1$ for FE and 2 for CN.

% !split
\subsection*{Exact numerical solution}

@@

Such an exact discrete solution is unusual, but very handy for analysis.

% !split
\subsection*{Stability}

@@

\noindent
% !split
\subsection*{Computation of stability in this problem}

@@

\noindent
% !split
\subsection*{Explanation of problems with Forward Euler}

@@

\noindent
% !split
\subsection*{Explanation of problems with Crank-Nicolson}

@@

\noindent
% !split
\subsection*{Summary of stability}

@@

\noindent
% !split
\subsection*{Comparing amplification factors}

@@

A possible measure of accuracy: $\Aex - A$

% !split
\subsection*{Plot of amplification factors}

% inline figure
\centerline{\includegraphics[width=0.9\linewidth]{fig-analysis/A_factors.pdf}}

% !split
\subsection*{$p=a\Delta t$ is the important parameter for numerical performance}

@@

\noindent

% --- begin paragraph admon ---
\paragraph{Another ``proof'' why $p=a\Delta t$ is key.}
If we scale the model
by $\bar t=at$, $\bar u=u/I$, we get
$d\bar u/d\bar t = -\bar u$, $\bar u(0)=1$ (no physical parameters!).
The analysis show that $\Delta \bar t$ is key, corresponding to
$a\Delta t$ in the unscaled model.
% --- end paragraph admon ---

% !split
\subsection*{Series expansion of amplification factors}

@@

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
>>> from sympy import *
>>> # Create p as a mathematical symbol with name 'p'
>>> p = Symbol('p')
>>> # Create a mathematical expression with p
>>> A_e = exp(-p)
>>>
>>> # Find the first 6 terms of the Taylor series of A_e
>>> A_e.series(p, 0, 6)
1 + (1/2)*p**2 - p - 1/6*p**3 - 1/120*p**5 + (1/24)*p**4 + O(p**6)

>>> theta = Symbol('theta')
>>> A = (1-(1-theta)*p)/(1+theta*p)
>>> FE = A_e.series(p, 0, 4) - A.subs(theta, 0).series(p, 0, 4)
>>> BE = A_e.series(p, 0, 4) - A.subs(theta, 1).series(p, 0, 4)
>>> half = Rational(1,2)  # exact fraction 1/2
>>> CN = A_e.series(p, 0, 4) - A.subs(theta, half).series(p, 0, 4)
>>> FE
(1/2)*p**2 - 1/6*p**3 + O(p**4)
>>> BE
-1/2*p**2 + (5/6)*p**3 + O(p**4)
>>> CN
(1/12)*p**3 + O(p**4)
\end{lstlisting}\end{cod}
\noindent

% !split
\subsection*{Error in amplification factors}

@@

\begin{equation}
A-\Aex = \left\lbrace\begin{array}{ll}
\Oof{\Delta t^2}, & \hbox{Forward and Backward Euler},\\ 
\Oof{\Delta t^3}, & \hbox{Crank-Nicolson}
\end{array}\right.
\end{equation}

% !split
\subsection*{The fraction of numerical and exact amplification factors}

@@

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
>>> FE = 1 - (A.subs(theta, 0)/A_e).series(p, 0, 4)
>>> BE = 1 - (A.subs(theta, 1)/A_e).series(p, 0, 4)
>>> CN = 1 - (A.subs(theta, half)/A_e).series(p, 0, 4)
>>> FE
(1/2)*p**2 + (1/3)*p**3 + O(p**4)
>>> BE
-1/2*p**2 + (1/3)*p**3 + O(p**4)
>>> CN
(1/12)*p**3 + O(p**4)
\end{lstlisting}\end{cod}
\noindent
Same leading-order terms as for the error measure $A-\Aex$.

% !split
\subsection*{The true/global error at a point}
\label{decay:analysis:gobal:error}

@@

\noindent
% !split
\subsection*{Computing the global error at a point}

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
>>> n = Symbol('n')
>>> u_e = exp(-p*n)   # I=1
>>> u_n = A**n        # I=1
>>> FE = u_e.series(p, 0, 4) - u_n.subs(theta, 0).series(p, 0, 4)
>>> BE = u_e.series(p, 0, 4) - u_n.subs(theta, 1).series(p, 0, 4)
>>> CN = u_e.series(p, 0, 4) - u_n.subs(theta, half).series(p, 0, 4)
>>> FE
(1/2)*n*p**2 - 1/2*n**2*p**3 + (1/3)*n*p**3 + O(p**4)
>>> BE
(1/2)*n**2*p**3 - 1/2*n*p**2 + (1/3)*n*p**3 + O(p**4)
>>> CN
(1/12)*n*p**3 + O(p**4)
\end{lstlisting}\end{cod}
\noindent
Substitute $n$ by $t/\Delta t$:

@@

\noindent
% !split
\subsection*{Convergence}

The numerical scheme is convergent if the global error
$e^n\rightarrow 0$ as $\Delta t\rightarrow 0$.
If the error has a leading order term $\Delta t^r$, the
convergence rate is of order $r$.

% !split
\subsection*{Integrated errors}

@@

% --- begin paragraph admon ---
\paragraph{Summary of errors.}
Analysis of both the pointwise and the time-integrated true errors:

@@

\noindent
% --- end paragraph admon ---

% !split
\subsection*{Truncation error}

@@

\begin{equation}
\frac{\uex(t_{n+1})-\uex(t_n)}{\Delta t} + a\uex(t_n) = R^n \neq 0
\label{decay:analysis:trunc:Req}
\end{equation}

% !split
\subsection*{Computation of the truncation error}

@@

\[ R^n \approx \half\uex''(t_n)\Delta t\]
This is a mathematical expression for the truncation error.

% !split
\subsection*{The truncation error for other schemes}

@@

\[ R^{n+\half} \approx \frac{1}{24}\uex'''(t_{n+\half})\Delta t^2\]

% !split
\subsection*{Consistency, stability, and convergence}

@@

(Consistency and stability is in most problems
much easier to establish than
convergence.)

% ------------------- end of main content ---------------

% #ifdef PREAMBLE
\cleardoublepage\phantomsection  % trick to get correct link to Index
\printindex

\end{document}
% #endif
